```{r echo = FALSE}
library(readxl)
library(xtable)
library(tidyverse)
library(pander)
library(knitr)
```


# Literature Review

The resilience and connectivity of transport networks are a long-studied topic within transportation engineering in both theoretical and practical contexts. Within this long history however, there is variability in how scholars define resiliency. There are three basic definitions that researchers have used:

  - Resilience through Resistance: Resilient transportation networks have few and
  manageable vulnerabilities. This is typically addressed through robust
  facility-level engineering and risk management (e.g., Bradley, 2007). 
  
  - Resilience through Recovery: Resilient transportation networks are able to be
  repaired and returned to normal service without inordinate delay. This is
  accomplished through effective resource allocation and incident management
  (e.g., Zhang and Wang, 2016). 

  - Resilience through Operability in Crisis: Resilient transport networks are
  able to operate effectively with damaged or unusable links. It is this
  definition that is most relevant in the context of this study.

These definitions are not entirely mutually exclusive, and many researchers apply more than one definition in their work. The motivation of this research is to identify systemically critical facilities, and therefore we primarily consider literature using the third definition. We begin this review first by examining the study conducted by AEM on behalf of UDOT to identify vulnerable sections on the I-15 corridor. We then consider previous attempts in the academic literature to evaluate real and fabricated transportation networks. 

## Identifying Critical Links on I-15
AEM worked with UDOT to put together a team of individuals to develop an I-15 Corridor Risk and Resilience (R&R) Pilot in 2017. This project had a seven-step plan to understand the impact of physical threats to the Utah transportation network, specifically looking at two sections along I-15. These steps include asset characterization, threat characterization, consequence analysis, vulnerability assessment, threat assessment, risk/resilience assessment, and risk/resilience management. From these steps AEM was able to provide recommendations to UDOT that would improve resiliency along the corridor based on the criticality of each segment at risk. 

AEM's analysis consisted of: gathering characteristic data for each asset (length, width, depth, condition, etc.), determining a replacement cost for each asset, establishing an estimated service life for each asset, estimating (if not known) the design standard for each asset, establishing which magnitudes of each threat were to be analyzed, and gathering information on the likelihood of occurrence of each magnitude of each threat. These steps are further described in their paper using multiple different threat locations. The AEM R&R report provides a good template going forward for identifying links at risk, following the first definition of a resilient transportation network. The report also attempts to identify which links at risk are most critical, assessing a “criticality” score to the network based on the five data elements and categories given in Table \@ref(tab:criticality). A key observation of these criticality scores is that they do not accommodate alternate routes that the traffic could use were the link to become unavailable. In identifying the systemic resiliency of highway facilities – as implied by the third definition of resiliency – requires considering these alternate routes.

```{r criticality, echo = FALSE}
tibble(
  Criteria = c("AADT", "Truck AADT", "AASHTO Roadway Classification", 
               "Tourism Traffic", "Maintenance Crew")
) %>%
  knitr::kable(caption = "AEM Criticality Score Assignment")

```

### Attempts to Evaluate Systemic Resiliency
A number of researchers have conducted studies where they construct real or fabricated transportation networks, eliminate or degrade links in the network, and evaluate the changes the loss of these links introduced into some measure of network performance. Berdica (2002) attempts to identify, define and conceptualize vulnerability by envisioning analyses conducted with several vulnerability performance measures including travel time, delay, congestion, serviceability and accessibility. Here, Berdica defines accessibility as the ability for users to travel between origins and destinations for any number of reasons. She then uses the performance measures to define vulnerability as the level of reduced accessibility due to unfavorable operating conditions on the network. In particular, Berdica identifies a need for further research toward developing a framework capable of investigating reliability of transportation networks. 

In this section we will examine several attempts by numerous researchers to do precisely this using different measures of network performance. A consolidation of this discussion is summarized in Table 2, namely the methods that different researchers have used in examining network performance under duress. The measures can be consolidated in to three basic families:
A. Network Connectivity considers how isolated nodes of a network become when links are damaged.
B. Travel Time analysis considers how removing or degrading a link affects the shortest cost paths between network points.
C. Accessibility analysis considers how the population using the network has their daily patterns affected by the damaged network.

We discuss studies within each family in turn.

```{r, echo = FALSE}
Table2 <- read_excel("Data/Table2.xlsx")
Table2 %>%
  pander(caption = "Table 2. Network Performance Measures in Resiliency Analysis")
```


## Network Connectivity
Graph theory is the mathematical study of networks of nodes connected by edges (links). Within this discipline are the related concepts of network vulnerability and connectivity that have been accessed by researchers. In these studies, researchers tend to define critical links as those that connect to many other nodes (directly or indirectly), or as links whose loss isolates a number of nodes from the rest of the network.

Abdel-Rahim et al., (2007) explores multi-layered graph theory to attempt to asses transportation network criticality. The researchers develop a weighting scheme based on interconnectedness between three layers of the transportation system: the power, communication, and physical road networks. This scheme is applied to and combined with data gathered using an Origin-destination matrix to determine criticality of road network facilities based on several proximity and link and node volume criteria. Abdel-Rahim et al. attempts to determine resilience through resistance by identifying a metric for examining the interrelatedness of the three systems, all of which are vital to proper function of the transportation network. Their methodology highlights some of the advantages to using a multi-layered graph approach for determining levels of resiliency of transportation networks.

Similarly, Ip and Wang (2011), attempt to combine graph theory with friability in order to determine criticality of links or hub cities. Friability is defined as the reduction of network capacity caused by removing a link or node. The methodology presented relies on the ability to determine the weighted sum of the resilience of all nodes based on the weighted average of connectedness with other city nodes in the network. The authors determine that the recovery of transportability between two cities largely depends on redundant links between nodes. Ip and Wang also comment that most traffic managers are more concerned with the friability of single links rather than the friability of multiple links or an entire system. The author's hypothesis closely matches resilience through operability because the methodology seeks to determine resilience of a network that already has damaged links and nodes. Their main contribution to the field of transportation research is the idea of friability, that a network can break down for any number of reasons, of which transportation networks generally appear most susceptible to attack or disaster.
Another researcher uses graph theory to approach the optimization of transportation systems. 

Zhang et al. (2015) investigates the role of network topology, meaning the layout of the network. The authors provide several examples of network topology types including hub and spoke, grid, and ring networks. After computing resilience indexes, or general resilience levels of each type of network topology, Zhang determines that metrics such as throughput, connectivity and average reciprocal distance increase with greater linage, however they decrease as networks become larger. This is likely because larger networks are sparser, having fewer node connections, and therefore are less redundant. Allowing for preparedness measures helped increase levels of resiliency. The authors greatest contribution appears to be increased understanding of the role network size and type plays in determining resiliency. Understanding how interconnectedness increases or decreases resiliency is vital to traffic managers.
Hackl and Adey (2019) show the abilities of a Networks in Networks (NIN) approach, where nodes in a network represent other networks. Their methodology allows multiple types of data including topology, paths, and origin-destination data to be loaded into the same graph style network. The researchers prepare an example NIN with four layers. The NIN is able to effectively demonstrate changes to traffic flow. Some are unable to occur due to interrupted edges on the network. This form of network representation is effective for simplifying highly connected networks. The contribution of NIN networks is highly interesting, and should be investigated further to help aid in the reduction of time needed to locate critical infrastructure on transportation networks.

## Changes in Travel Time
Graph theoretical approaches to resiliency largely consider whether the network nodes become isolated after links are degraded or removed. Though isolation is an important problem, transportation networks often have multiple redundant paths between nodes; however, some of these paths may be considerably longer. It is therefore useful to determine how the path cost changes even if the nodes involved are still reachable on the network.

Peeta et al. (2010) constructs a model to evaluate the most efficient allocation of highway maintenance resources prior to an earthquake with the potential to disrupt transport network connectivity. The authors create a simplified graph of Istanbul, Turkey with 30 attributed links and 25 nodes, with each link having a specified failure probability that could be mitigated with investment. Using this graph, the researchers evaluate what happened to travel times between the origin-destination pairs as links from this network were disabled using a Monte Carlo simulation. The authors showed that this problem is tractable with a locally optimal solution existing. 

Ibrahim (2011) provides an alternative heuristic approach for determining vulnerability of infrastructure by estimating the cost of single link failure based on the increase in shortest path travel time due to increased congestion levels. Ibrahim proposes a hybrid heuristic approach that calculates the traditional user-equilibrium assignment for finding the first set of costs, and then fixes those costs for all following iterations to determine the effects of failure on overall travel time of the system. Ibrahim’s novel heuristic approach is important because of its ability to drastically reduce computation time for larger networks while providing accurate results that closely match results found from traditional modeling methods.

Omer et al. (2013) proposes a methodology for assessing the resiliency of physical infrastructure during disruptions. To do this, the authors use a network model to build an origin-destination matrix that allows initial network loading and analysis. Omer’s model uses several metrics, but the main metric used to determine resiliency is the difference in travel time between a disturbed and undisturbed network. Omer’s framework is applied to an actual network between New York City and Boston for analysis. Changes in demand, travel time, mode choice and route choice are tracked for analysis. Omer’s framework supports operability of transportation networks due to the way it analyzes networks experiencing suboptimal circumstances. The authors work identifies key parameters that should be measured to assess resiliency during disruptive events.

Jaller et al. (2015) seeks to identify critical infrastructure based on increased travel time, or reduced capacity due to disaster. The proposed methodology utilizes user-equilibrium to determine proper initial network loading. Then, the shortest path between one origin and one destination can be identified. To implement damage to the network, a link is cut, and then the next shortest path is found. This process is followed for all links in the system in order to determine a sense of the criticality of each link to network resiliency. The analysis is carried out for each O-D pair, and the nodes with greatest change in travel time are determined to be the most critical. Jaller’s methodology allows traffic managers to identify critical paths for mitigation purposes before the occurrence of disaster through careful analysis.

## Changes in Accessibility
Accessibility refers to the ease with which individuals can reach the destinations that matter to them; this is an abstract idea but one that has been quantified in numerous ways. Dong et al., (2006) provides a helpful framework for understanding various quantitative definitions of accessibility that we will simplify here. The most elementary definition of accessibility is whether a destination is within an isochrone, or certain distance. This measure is often represented as a count, e.g., the number of jobs reachable from a particular location within thirty minutes travel time by a particular mode. Mathematically,
$A_i=\sum_j X_j I_{ij}$ ,where $I_{ij} = 1$ if $d_{ij} < D$ and $I_{ij}=0$ if $d_{ij} > D $
where the accessibility A at point i is the sum of the all the jobs or other destinations X at other points j. $I_{ij}$ is an indicator function equal to zero if the distance between the points $d_{ij}$ is less than some asserted threshold (e.g., thirty minutes of travel time). By relaxing the assumption of a binary isochrone and instead using the distance directly, we can derive the so-called gravity model,
$A_i=\sum_j X_j f(d_{ij})$
where the function $f(d_{ij)}$ is often a negative exponential with a calibrated impedance coefficient. An extension of the gravity model is to use the logsum term of a multinomial logit destination choice model,
$A_i=ln\sum_j β_d d_{ij} + X_j β$
Where the parameters β are estimated from choice surveys or calibrated to observed data. 

The logsum term has numerous benefits outlined by Handy and Niemeier (1997) and Dong et al. (2007); namely, the measure is based in actual choice theory, and can include multiple destination types and travel times by multiple different modes. Accessibility measures of any kind are important in resiliency analysis because a damaged transport network will limit the ability of people to access the full variety of destinations they otherwise would.
Guers and van Wee (2004) provide a review of accessibility measures such as those above up to 2004. Of the papers they reviewed, Vickerman (1974), Ben-Akiva and Lerman (1979), Geurs and Ritsema van Eck (2001) used isochrone type methods, Stewart (1947), Hansen (1959), Ingram (1971), and Vickerman (1971), and Anas (1983) used gravity, and Neuburger (1971), Leonardi (1987), Williams and Senior (1978), Koenig( 1980), Anas (1983), Ben-Akiva and Lerman (1985), Sweet (1997), Niemier (1997), Handy and Niemier (1997), Levine (1998), and Miller (1999) used or suggested logsums. They highlight the importance of using person-based measures such as these in evaluating network vulnerability and resiliency.

Taylor (2008) also examines accessibility as a way to analyze network vulnerabilities, mainly through logsum analysis. Taylor begins by explaining the concepts of vulnerability and accessibility, which further help to explain their interrelatedness and key distinctions. He then applies logsum analysis to provide a framework for solving problems of accessibility. Taylor’s method for identifying vulnerabilities involves modeling travel demand, network topology, capacity, and road geometry in a manner that closely resembles a graph network. He does make one key distinction between links and nodes, however. A Node is vulnerable if link loss to that node diminishes its accessibility, while a Link is critical if its loss significantly diminishes accessibility within the network. One drawback to Taylor’s methodology at this point, is that it is intended to be used after critical locations have been identified.

Ganin et al. (2017) attempts to investigate resiliency through a disruption of 5% of the roadways in 40 urban networks within the United States. The employed methodology determines that Salt Lake City has the most resilient transportation network while Washington D.C. has the least resilient. This determination is based on a comparison of the network after links are damaged versus before. The authors work three factors into each model, namely the proportionality coefficient, minimum speed, and vehicle size adjustment factors. These essentially account for differences in car-truck ratios, average speed, and average vehicle length. Using a gravity model—a model which accounts for attraction and distance between zones—the authors were able to estimate the average annual delay per commuter. They used this to determine network efficiency. Ganin noted that traffic delay times increase significantly as road segments are broken. The main benefit from this article is the use of a gravity model to determine ways in which accurate transportation models can be calculated.

## Summary
The lessons learned from the events in Minneapolis and Atlanta demonstrate that when transportation networks are damaged or degraded by link failure, multiple changes result. Traffic diverts to other facilities and other modes, and some people make fundamental changes to their daily activity patterns, choosing new destinations or eliminating trips entirely. Numerous other researchers have identified methodologies to capture the effects, or at least the costs, of these potential changes in modeled crisis events. We are able to learn from these methodologies to create a methodology on a state-wide level.